---
title: "TSA final project"
output:
  pdf_document: default
  html_document: default
date: "2026-04-12"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(lubridate)
library(ggplot2)
library(forecast)  
library(tseries)
library(tidyverse)
library(smooth)
library(dplyr)

```

# Introduction, Motivation, Relevance, Objectives

# Introduction
Electricity demand forecasting is an essential foundation for power grid operation and planning, especially given the growing frequency of extreme weather events caused by climate change. The California Independent System Operator (CAISO), headquartered in Folsom, California, manages one of the largest electricity markets in the United States and serves over 32 million users within its regulated area. Each year, CAISO would deliver more than 260 million megawatt-hours of electricity (CAISO, n.d.). From 2014 to 2024, the annual peak demand of CAISO ranges between 43,789 and 51,479 MW (CAISO, 2025), but during the September 2022 extreme heat wave, it surged to an all-time record of 52,061 MW, releasing an emergency energy alert and affecting over 500,000 customers for potential outage concerns (Kavya Balaraman, 2022).

This record-breaking load not only requires greater precision and accuracy in real-time grid balancing but also poses significant challenges for demand forecasting models. Because commonly used linear models may fail to adequately capture the nonlinear relationship between temperature and demand, and this underestimation could be expanded during heat wave periods (Wang et al., 2021). Therefore, motivated by this concern, our final project analyzes how different time-series models can forecast electricity demand in CAISO using historical temperature data from California, especially during heatwave periods. Specifically, we adopt the SARIMA modeling framework as our baseline and progressively incorporate temperature and an extreme heat indicator as exogenous variables. This incremental design helps us to systematically evaluate how precisely each variable contributes to forecast accuracy.

# Motivation
Since the 1900s, the average temperature in California has risen by around 2.5 to 3 degrees, and the frequency of heat waves in the United States has doubled since the 1980s (Klivans, 2024). At the same time, compared to the 1960s, in 2024, the average length of the heat wave seasons has increased by 46 days (Center for Climate And Energy Solutions, 2025). Therefore, the significantly increasing extreme heat waves pose a direct threat to the reliability of the power grid.

In recent years, CAISO has been significantly affected by extreme heat, and among these impacts, three major events illustrate the scale of the impact.

First, in August 2020, CAISO experienced several rolling blackouts, with around 492,000 customers losing power for 15 to 150 minutes on August 14, followed by another 321,000 customers losing power for 8 to 90 minutes on August 15 (California ISO, 2020). These rotating outages were mainly caused by improper planning and preparation, early-evening supply gaps during solar ramp-down, and CAISO energy trading market issues related to convergence bidding (California ISO, 2020).

Second, in September 2022, heat waves pushed CAISO peak load to an all-time record of 52,061 MW (California ISO, 2022), and the demand across the entire Western Interconnection hit 167,500 MW (California ISO, 2024). In response, CAISO issued Flex Alerts for 10 consecutive days and triggered a Stage 3 Energy Emergency Alert to avoid rolling blackouts (EIA, 2023). According to EIA (2022), natural gas surged to 60% of the fuel mix, up from around 30% under normal conditions, to meet this electricity demand.

Lastly, on July 11, 2024, CAISO demand reached a peak of 43,969 MW, and the heat wave brought over two weeks of triple-digit temperatures (100°F+). Because CAISO added 11,000 MW of clean energy resources and battery storage to the grid, it successfully managed the event without issuing any Flex Alerts or emergency alerts (Subakti, 2024).

Together, these three events demonstrate the importance of accurate demand forecasting for supporting resource planning and operational decisions under extreme heat waves.

# Data Set Description

```{r}
#Importing time series data from text file
load_temp_data <- read.csv(
  file="./caiso_load_temp_final.csv",
  header=TRUE)

head(load_temp_data)

load_temp_data$date <- as.Date(load_temp_data$date)
load_temp_data <- load_temp_data %>% arrange(date)
```


# Methods
Dataset split: 
Prior to being utilized for modeling purposes, the data set was divided into two separate time frames: A training time frame (2021-2024); and a validation time frame (2025). 

Variable Selection: 
We selected daily peak load as the target variable because electricity system planning and capacity decisions are driven by peak demand rather than average demand. This reflects system stress during high-demand periods, which is more relevant for grid reliability.

Temperature (tmax_mean) is included as the primary exogenous variable, as it directly represents peak daily heat exposure and that leads to high electricity demand for cooling.

To further capture potential nonlinear effects caused by occasional heatwave events, we constructed a dummy variable as an extreme heat indicator using the 95th percentile of temperature (~92°F).

Model 1: SARIMA (Baseline Model)
We utilized a SARIMA model as our baseline model, using only historic electricity demand data to identify autoregressive structure within the data and weekly seasonality, 

Model 2: SARIMAX (with Temperature)

Model 2 includes daily mean maximum temperature as an additional explanatory variable to the SARIMA model. The inclusion of temperature as a predictor (xreg) will enable us to model the continuously changing relationship between temperature and electric energy usage. 

Model 3: SARIMAX (with Temperature & Extreme Heat)
Both temperature and extreme heat indicator were included as predictors in model 3. This  allows us to assess both continuous changes in electric energy consumption resulting from temperature changes along with possible discrete increases in electric energy consumption resulting from extreme heat events.

Model Assessment
Each model's prediction accuracy was measured using RMSE, MAE, and MAPE. The use of these three measurements allowed us to compare both the absolute performance of each model as well as their relative performance.


# Result
## Model 1 (Baseline): SARIMA

```{r}
# Split dataset
train_data <- load_temp_data %>%
  filter(date >= as.Date("2021-01-01") & date <= as.Date("2024-12-31"))

valid_data <- load_temp_data %>%
  filter(date >= as.Date("2025-01-01") & date <= as.Date("2025-12-31"))

train_ts <- ts(train_data$daily_peak_load, frequency = 7)

valid_ts <- ts(
  valid_data$daily_peak_load,
  start = end(train_ts) + c(0, 1),
  frequency = frequency(train_ts)
)
```

```{r}
fit_sarima <- auto.arima(train_ts, seasonal = TRUE)
summary(fit_sarima)

# validation 
forecast_2025 <- forecast(fit_sarima, h = length(valid_ts))
accuracy(forecast_2025, as.numeric(valid_data$daily_peak_load))

autoplot(forecast_2025) +
  autolayer(ts(valid_data$daily_peak_load,
               start = end(train_ts) + c(0,1),
               frequency = frequency(train_ts)),
            series = "Actual",
            color = "red") +
  xlab("Date") +
  ylab("Daily Peak Load") +
  ggtitle("SARIMA (Baseline): Forecast vs Actual (2025)")

```

```{r}
plot_df_sarima <- valid_data %>%
  mutate(forecast = as.numeric(forecast_2025$mean))

ggplot(plot_df_sarima, aes(x = date)) +
  geom_line(aes(y = daily_peak_load, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = forecast, color = "Forecast"), linewidth = 0.8) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  labs(
    title = "Baseline SARIMA: Forecast vs Actual (2025)",
    x = "Date",
    y = "Daily Peak Load",
    color = NULL
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Model 2: SARIMAX

```{r}
xreg_train <- as.matrix(train_data$tmax_mean)
xreg_valid <- as.matrix(valid_data$tmax_mean)

sum(is.na(xreg_train))
sum(is.na(xreg_valid))

# Fit model
fit_sarimax <- auto.arima(train_ts,
                          xreg = xreg_train,
                          seasonal = TRUE)
summary(fit_sarimax)

forecast_sarimax_2025 <- forecast(fit_sarimax,
                            xreg = xreg_valid,
                            h = nrow(valid_data))
accuracy(forecast_sarimax_2025, as.numeric(valid_data$daily_peak_load))

autoplot(forecast_sarimax_2025) +
  autolayer(ts(valid_data$daily_peak_load,
               start = end(train_ts) + c(0,1),
               frequency = frequency(train_ts)),
            series = "Actual",
            color = "red") +
  xlab("Date") +
  ylab("Daily Peak Load") +
  ggtitle("SARIMAX: Forecast vs Actual (2025)")



```

```{r}
plot_df_sarimax <- valid_data %>%
  mutate(forecast = as.numeric(forecast_sarimax_2025$mean))

ggplot(plot_df_sarimax, aes(x = date)) +
  geom_line(aes(y = daily_peak_load, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = forecast, color = "Forecast"), linewidth = 0.8) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  labs(
    title = "SARIMAX: Forecast vs Actual (2025)",
    x = "Date",
    y = "Daily Peak Load",
    color = NULL
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Model 3: Sarimax + Extreme heat dummy

```{r}
# Set the percentile of extreme heat
heat_threshold <- quantile(load_temp_data$tmax_mean, 0.95, na.rm = TRUE)

load_temp_data$extreme_heat <- ifelse(
  load_temp_data$tmax_mean >= heat_threshold, 1, 0
)

table(load_temp_data$extreme_heat)
heat_threshold

train_data <- load_temp_data %>%
  filter(date >= as.Date("2021-01-01") & date <= as.Date("2024-12-31"))

valid_data <- load_temp_data %>%
  filter(date >= as.Date("2025-01-01") & date <= as.Date("2025-12-31"))

train_ts <- ts(train_data$daily_peak_load, frequency = 7)

xreg_train_m3 <- cbind(
  tmax = train_data$tmax_mean,
  extreme = train_data$extreme_heat
)

xreg_valid_m3 <- cbind(
  tmax = valid_data$tmax_mean,
  extreme = valid_data$extreme_heat
)

fit_sarimax_m3 <- auto.arima(
  train_ts,
  xreg = xreg_train_m3,
  seasonal = TRUE
)

summary(fit_sarimax_m3)

forecast_sarimax_m3_2025 <- forecast(
  fit_sarimax_m3,
  xreg = xreg_valid_m3,
  h = nrow(valid_data)
)

accuracy(forecast_sarimax_m3_2025, as.numeric(valid_data$daily_peak_load))

autoplot(forecast_sarimax_m3_2025) +
  autolayer(ts(valid_data$daily_peak_load,
               start = end(train_ts) + c(0,1),
               frequency = frequency(train_ts)),
            series = "Actual",
            color = "red") +
  xlab("Date") +
  ylab("Daily Peak Load") +
  ggtitle("SARIMAX + extreme heat dummy: Forecast vs Actual (2025)")
```

```{r}
plot_df_sarimax_dummy <- valid_data %>%
  mutate(forecast = as.numeric(forecast_sarimax_m3_2025$mean))

ggplot(plot_df_sarimax_dummy, aes(x = date)) +
  geom_line(aes(y = daily_peak_load, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = forecast, color = "Forecast"), linewidth = 0.8) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  labs(
    title = "SARIMAX + Extreme Heat Dummy: Forecast vs Actual (2025)",
    x = "Date",
    y = "Daily Peak Load",
    color = NULL
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Model 4: TBATS

```{r}
# Create multi-seasonal ts with weekly + annual cycles
train_msts <- msts(train_data$daily_peak_load,
                   seasonal.periods = c(7, 365.25),
                   start = decimal_date(as.Date("2021-01-01")))

# Fit TBATS
fit_tbats <- tbats(train_msts)
summary(fit_tbats)

# Forecast 2025
forecast_tbats_2025 <- forecast(fit_tbats, h = nrow(valid_data))

# Accuracy
accuracy(forecast_tbats_2025, as.numeric(valid_data$daily_peak_load))

# plot
autoplot(forecast_tbats_2025) +
  autolayer(ts(valid_data$daily_peak_load,
               start = end(train_msts) + c(0, 1),
               frequency = 7),
            series = "Actual",
            color = "red") +
  xlab("Date") +
  ylab("Daily Peak Load") +
  ggtitle("TBATS (Multi-Seasonal): Forecast vs Actual (2025)")

plot_df_tbats <- valid_data %>%
  mutate(forecast = as.numeric(forecast_tbats_2025$mean))

ggplot(plot_df_tbats, aes(x = date)) +
  geom_line(aes(y = daily_peak_load, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = forecast, color = "Forecast"), linewidth = 0.8) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  labs(
    title = "TBATS (Multi-Seasonal): Forecast vs Actual (2025)",
    x = "Date",
    y = "Daily Peak Load",
    color = NULL
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```

## Model 5: Dynamic Harmonic Regression (ARIMA + Fourier Terms)

```{r}
# Step 1: Multi-seasonal ts
train_msts <- msts(train_data$daily_peak_load,
                   seasonal.periods = c(7, 365.25),
                   start = decimal_date(as.Date("2021-01-01")))

# Step 2: Find optimal K via AICc (weekly K1=1:3, annual K2=1:5)
best_k    <- c(1, 1)
best_aicc <- Inf

for (k1 in 1:3) {
  for (k2 in 1:5) {
    xreg_k <- cbind(
      fourier(train_msts, K = c(k1, k2)),
      tmax = train_data$tmax_mean
    )
    fit_k <- auto.arima(train_msts,
                        xreg      = xreg_k,
                        seasonal  = FALSE,
                        stepwise  = TRUE,
                        approximation = TRUE)  # faster search
    cat("K =", k1, k2, "| AICc =", fit_k$aicc, "\n")
    if (fit_k$aicc < best_aicc) {
      best_aicc <- fit_k$aicc
      best_k    <- c(k1, k2)
    }
  }
}

cat("\nBest K:", best_k, "\n")

# Step 3: Fit final model with best K + tmax
fourier_train <- cbind(
  fourier(train_msts, K = best_k),
  tmax = train_data$tmax_mean
)

fourier_valid <- cbind(
  fourier(train_msts, K = best_k, h = nrow(valid_data)),
  tmax = valid_data$tmax_mean
)

fit_dhr <- auto.arima(train_msts,
                      xreg      = fourier_train,
                      seasonal  = FALSE,
                      stepwise  = FALSE,
                      approximation = FALSE)

summary(fit_dhr)

# Step 4: Forecast 2025
forecast_dhr_2025 <- forecast(fit_dhr,
                              xreg = fourier_valid,
                              h    = nrow(valid_data))

accuracy(forecast_dhr_2025, as.numeric(valid_data$daily_peak_load))

# Step 5: Plot
plot_df_dhr <- valid_data %>%
  mutate(forecast = as.numeric(forecast_dhr_2025$mean))

ggplot(plot_df_dhr, aes(x = date)) +
  geom_line(aes(y = daily_peak_load, color = "Actual"), linewidth = 0.8) +
  geom_line(aes(y = forecast, color = "Forecast"), linewidth = 0.8) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  labs(
    title = paste0("DHR (K=", best_k[1], ",", best_k[2], ") + tmax: Forecast vs Actual (2025)"),
    x = "Date",
    y = "Daily Peak Load",
    color = NULL
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Model comparison
```{r}
rmse_sarima    <- accuracy(forecast_2025,
                           as.numeric(valid_data$daily_peak_load))[2, "RMSE"]

rmse_sarimax   <- accuracy(forecast_sarimax_2025,
                           as.numeric(valid_data$daily_peak_load))[2, "RMSE"]

rmse_sarimax_d <- accuracy(forecast_sarimax_m3_2025,
                           as.numeric(valid_data$daily_peak_load))[2, "RMSE"]

rmse_tbats     <- accuracy(forecast_tbats_2025,
                           as.numeric(valid_data$daily_peak_load))[2, "RMSE"]

rmse_dhr       <- accuracy(forecast_dhr_2025,
                           as.numeric(valid_data$daily_peak_load))[2, "RMSE"]

# Build comparison dataframe
comparison_df <- data.frame(
  Model = c("SARIMA (Baseline)",
            "SARIMAX",
            "SARIMAX + Heat Dummy",
            "TBATS",
            "DHR + tmax"),
  RMSE = round(c(rmse_sarima,
                 rmse_sarimax,
                 rmse_sarimax_d,
                 rmse_tbats,
                 rmse_dhr), 2)
) %>%
  arrange(RMSE) %>%
  mutate(Rank = row_number())

# Print table
print(comparison_df)

# Bar chart
ggplot(comparison_df, aes(x = reorder(Model, RMSE), y = RMSE, fill = RMSE)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = format(round(RMSE, 1), big.mark = ",")),
            hjust = -0.1, size = 3.8, fontface = "bold") +
  coord_flip(clip = "off") +
  scale_fill_gradient(low = "#2c7bb6", high = "#d7191c") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)),
                     labels = scales::comma) +
  labs(
    title = "Model Comparison: RMSE on 2025 Validation Set",
    subtitle = "Lower RMSE = better forecast accuracy",
    x = NULL,
    y = "RMSE"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey50", size = 10),
    axis.text.y   = element_text(size = 11),
    panel.grid.major.y = element_blank()
  )
```
# Results
The results for 2025 show that incorporating temperature significantly improves forecasting accuracy. The SARIMAX model reduces RMSE from 4245.6 to 3014.7 (approximately 29% improvement), with similar improvements in MAE and MAPE. This confirms that electricity demand in CAISO is strongly driven by temperature, as higher temperatures increase cooling demand and lead to higher electricity load.

However, adding the extreme heat dummy provides only marginal additional improvement (RMSE: 3014.7 → 2996.0). This suggests that the continuous temperature variable already captures most of the variation in demand, including during heatwave periods. Overall, the results indicate that electricity demand responds smoothly to temperature changes rather than rare extreme heatwave events.


# Summary and Conclusions


# Reference

CAISO. (n.d.). How electricity flows | California ISO. Caiso.com. Retrieved April 24, 2026, from https://www.caiso.com/about/our-business/how-electricity-flows

CAISO. (2025). 2025 Summer Loads and Resources Assessment. Caiso.com. https://www.caiso.com/content/summer-loads-resources-assessment/2025/index.html

California ISO. (2020). This page intentionally left blank. https://www.caiso.com/documents/preliminary-root-cause-analysis-rotating-outages-august-2020.pdf

California ISO. (2022). NEWS RELEASE California ISO posts analysis of September heat wave. https://www.caiso.com/Documents/california-iso-posts-analysis-of-september-heat-wave.pdf

California ISO. (2024). Summer Market Performance Report. https://www.caiso.com/documents/summer-market-performance-report-july-2024.pdf

Center for Climate And Energy Solutions. (2025). Heat Waves and Climate Change. Center for Climate and Energy Solutions; C2ES. https://www.c2es.org/content/heat-waves-and-climate-change/

EIA. (2022). California fuel mix changes in response to September heat wave. Www.eia.gov. https://www.eia.gov/todayinenergy/detail.php?id=53939

EIA. (2023). California consumers respond to appeals for electricity conservation during heat wave - U.S. Energy Information Administration (EIA). Eia.gov. https://www.eia.gov/todayinenergy/detail.php?id=54039

Kavya Balaraman. (2022, September 7). California ISO narrowly avoids rolling outages as peak demand hits record 52 GW. Utility Dive. https://www.utilitydive.com/news/california-iso-narrowly-avoids-outages-but-warns-as-peak-demand-hits-record/631280/

Klivans, L. (2024, October 4). Yes, These California Heat Waves Are Connected to Climate Change. Here’s How. Kqed.org. https://www.kqed.org/science/1994643/yes-these-california-heat-waves-are-connected-to-climate-change-heres-how

Subakti, D. (2024, July 15). Managing the July 2024 heat wave with our partners in California and the West | California ISO. Caiso.com. https://www.caiso.com/about/news/energy-matters-blog/managing-the-july-2024-heat-wave-with-our-partners-in-california-and-the-west

Wang, Z., Hong, T., Li, H., & Ann Piette, M. (2021). Predicting city-scale daily electricity consumption using data-driven models. Advances in Applied Energy, 2, 100025. https://doi.org/10.1016/j.adapen.2021.100025
